---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### Loading and preprocessing the data
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)

options(scipen = 1, digits = 2)

dataDir <- "./data"
activityFile <- paste(dataDir,"/repdata_data_activity.zip", sep="");

if (!file.exists(dataDir)) {
  dir.create(file.path(dataDir))
}

if (!file.exists(activityFile)) {
  activityFile <- paste(dataDir,"/repdata_data_activity.zip", sep="");
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", activityFile)
}

data <- read.csv(unz(activityFile,"activity.csv"))
data <- mutate(data, daytype = ifelse(wday(date, label=T) %in% c("Sat","Sun"),'weekend','weekday'))
data <- mutate(data, dow = wday(date, label=T))

totStepsPerDay <- data[!is.na(data$steps),] %>%
  group_by(date) %>% 
  summarise( totSteps = sum(steps))


medianStepsPerDay <- round(median(totStepsPerDay$totSteps),0)
meanStepsPerDay <- round(mean(totStepsPerDay$totSteps),0)

ggplot(data=totStepsPerDay, aes(x=totSteps)) + 
                geom_histogram()
```

### What is mean total number of steps taken per day?

1. Median steps per day: `r medianStepsPerDay`  
2. Mean steps per day: `r meanStepsPerDay`

### What is the average daily activity pattern?

```{r}
meanByInterval <- data[!is.na(data$steps),] %>%
  group_by(interval) %>%
  summarise_each_(funs(mean), vars = c("interval","steps"))

with(meanByInterval,plot(interval, steps, type="l", ylab="Steps", xlab = "Interval"))

```

**Interval with Maximum average steps**: `r meanByInterval[meanByInterval$steps == max(meanByInterval$steps),]$interval` (`r round(max(meanByInterval$steps))` average steps)

### Imputing missing values
```{r}
numNA <- nrow(data[is.na(data$steps),])
```

1. Number of step observations with NA: `r numNA`

2. Fill in NA values using mean of the interval (by the day of week)

```{r}

mid <- data[!is.na(data$steps),] %>%
  group_by(interval, dow) %>%
  summarise_each_(funs(mean), vars = c("interval","steps","dow"))

data$StepsIntDOW <- as.numeric(apply(data,1, function(x) if( is.na(x[1])) round(mid[mid$interval == as.numeric(x[3]) & mid$dow ==  x[5],]$steps,0) else x[1][[1]][[1]]))

```

Histogram of steps per day with NAs filled in:

```{r}
totStepsPerDayIntDOW <- data %>%
  group_by(date) %>% 
  summarise( totSteps = sum(StepsIntDOW))

ggplot(data=totStepsPerDayIntDOW, aes(x=totSteps)) + 
  geom_histogram()

```

Calculate/display mean and median steps per day with NAs filled in

```{r}
totStepsPerDayIntDOW <- data %>%
  group_by(date) %>% 
  summarise( totSteps = sum(StepsIntDOW))

medianStepsPerDayIntDOW <- median(totStepsPerDayIntDOW$totSteps)
meanStepsPerDayIntDOW <- mean(totStepsPerDayIntDOW$totSteps)
#nNA <- nrow(data[is.na(data$steps),])
```
1. Mean steps per day: `r meanStepsPerDayIntDOW`
2. Median steps pers day: `r medianStepsPerDayIntDOW`

### Are there differences in activity patterns between weekdays and weekends?

```{r}
pRAW <- ggplot(data=data, aes(x=interval, y=steps)) + 
          geom_line() + 
          facet_grid(daytype ~ .) +
          ggtitle("Weekend vs Weekday Activity \n (NA not removed)") + 
          ylab("Steps") + 
          xlab("Interval")

pDOW <- ggplot(data=data, aes(x=interval, y=StepsIntDOW)) + 
          geom_line() + 
          facet_grid(daytype ~ .) +
          ggtitle("Weekend vs Weekday Activity \n (NA replaced with Mean of steps by Interval, Day of Week)") + 
          ylab("Steps") + 
          xlab("Interval")

par(mfrow=c(1,2))
pRAW
pDOW
```